---
title: "**CRIME DATA ANALYSIS**"
output: html_document
---

### **MAIN OBJECTIVE**

The objective of this notebook is to explore the crime dataset found in kaggle at: <https://www.kaggle.com/datasets/haseefalam/crime-dataset/data?select=Crime_Data_from_2020_to_Present.csv>.

### **NECESSARY LIBRARIES**

Those are the libraries that are going to be used across this notebook.

```{r}
library(ggplot2)
library(lubridate)
```

### **DATA LOAD**

```{r}
setwd("C://Users//riosa//documents//datasets//crime_dataset") # sets the directory to the specified path
crimeset <- read.csv("Crime_Data_from_2020_to_Present.csv") # loads the dataset
```

### **EXPLORATORY DATA ANALYSIS**

```{r}
dim(crimeset)
```
The dataset counts with 955339 observations and 28 variables.
Let's check the variable names.
```{r}
names(crimeset)
```
There are missing values in some variable?
```{r}
missing_table <- as.matrix(colSums(apply(X = crimeset, MARGIN = 2, FUN = is.na)))
missing_table
```
The percentages of missing values in the different variables are:
```{r}
for (i in 1:dim(missing_table)[1]) {
  if (missing_table[i, 1] != 0) {
    cat("Percentage of missings of variable ", names(missing_table[i, ]), ":", round(100*missing_table[i, 1]/955339, 2), "%\n")
  }
}
```
Variables Crm.Cd.2, Crm.Cd.3 and Crm.Cd.4 have more than a 90 % of missing values, so they are hardly useful for modelling.

Let's explore now the contents of each variable:

#### **1) Variable DR_NO:**

```{r}
head(crimeset$DR_NO)
length(unique(crimeset$DR_NO))
length(unique(crimeset$DR_NO)) == dim(crimeset)[1]
```
This variable contains a unique identifier for each observation, so this variable will not be useful for modelling.

#### **2) Variable Date.Rptd:**

```{r}
head(crimeset$Date.Rptd)
```
This variable contains dates and times. The first thing we will do is to check how many different hours the variable takes.
```{r}
# Let's see how to proceed for one case...
paste(unlist(strsplit(crimeset$Date.Rptd[1], " "))[2:3], collapse = " ")

# Let's create an auxiliar funciton to perform the above task
getHour <- function(value) {
  # Input a character string like "03/01/2020 12:00:00 AM"
  # Return a character string like "12:00:00 AM"
  # Input is assumed to be passed correctly by the user
  return(paste(unlist(strsplit(value, " "))[2:3], collapse = " "))
}

# Testing...
getHour(crimeset$Date.Rptd[1])
getHour(crimeset$Date.Rptd[2])
getHour(crimeset$Date.Rptd[3])

# Let's store all the hours in a vector
checkHours <- vapply(crimeset$Date.Rptd, getHour, character(1))
# Vector obtained
head(checkHours)
# Let's check how many different values it takes...
length(unique(checkHours))
```
There is only one hour that repeats across all the rows for this variable, 12:00:00 AM, so let's remove them and only remain with the date. For doing this, let's add a new variable to the dataset.
```{r}
# Let's create an auxiliar function to do the process...
getDate <- function(value) {
  # Input: a character string with the format: "03/01/2020 12:00:00 AM"
  # Return: a character type with the format: "03/01/2020"
  # Input is assumed to be correctly passed by the user
  return(unlist(strsplit(value, " "))[1])
}

# Now let's create a new variable withe the dates in the dataset crimeset
crimeset$rptd_date <-  vapply(crimeset$Date.Rptd, getDate, character(1))
crimeset$rptd_date <- as.Date(crimeset$rptd_date, format = "%m/%d/%Y")

# Result
head(crimeset$rptd_date)
range(crimeset$rptd_date)
```
The range of the dates goes from the first of January of 2020 to the 24th of June of 2024.
```{r}
length(unique(crimeset$rptd_date))
```
There are only 1637 different dates for 955339 observations. Let's see which ones repeat the most.
```{r}
# Ten most repeated dates
head(sort(table(crimeset$rptd_date), decreasing = T), 10)
```
Let's check now the frequencies of each year.
```{r}
# Frequencies of the years
table(year(crimeset$rptd_date))/sum(table(year(crimeset$rptd_date)))
```
The year with more observations is 2022 with the 24.61 %, followed by 2023 with the 24.56 %. The year with less observations is 2024 which only accounts for the 8.86 % of the observations, although there are observations until the mid of 2024.

#### **3) Variable DATE.OCC:**

```{r}
head(crimeset$DATE.OCC)
```
Let's proceed like in the previous variable, first let's check if all the observations take
the same hour.
```{r}
# Let's store all the hours in a vector
checkHours2 <- vapply(crimeset$DATE.OCC, getHour, character(1))
head(checkHours2)
length(unique(checkHours2))
```
Again, there is the same hour for all the rows, so let's save in a new variable only the dates.
```{r}
crimeset$date_occ <-  vapply(crimeset$DATE.OCC, getDate, character(1))
crimeset$date_occ <- as.Date(crimeset$date_occ, format = "%m/%d/%Y")
head(crimeset$date_occ)
range(crimeset$date_occ)
```
The range of dates is the same of the previous variable.
```{r}
head(crimeset$rptd_date)
head(crimeset$date_occ)
```
It seems that the dates of the variable rptd_date can be always older or equeal than the dates of the variable date_occ, let's check it.
```{r}
checkPositive <- crimeset$rptd_date - crimeset$date_occ
head(checkPositive, 10)
length(which(checkPositive >= 0))
```
Confirmed, it's true for all rows.

#### **4) Variable TIME.OCC:**

```{r}
head(crimeset$TIME.OCC)
summary(crimeset$TIME.OCC)
head(sort(table(crimeset$TIME.OCC), decreasing = T), 10)
```
It's not clear the meaning and the format of this variable, by its name, it's supposed to contain hours and maybe minutes, but its format is of integer values from 1 to 2359. Values are repeated many times so maybe they can not take any value from 1 to 2359. It could be that 1 means 00:01 and 2359 means 23:59. For checking if this is possible let's test that in all the obervations with four digits the first two never take values bigger than 23, and that the last two digits never take values bigger than 59.
```{r}
# Let's check first only the observations of four digits.
checkFirstTwo <- crimeset$TIME.OCC[which(crimeset$TIME.OCC >= 1000)]%/%100
checkLastTwo <- crimeset$TIME.OCC[which(crimeset$TIME.OCC >= 1000)]%%100
range(checkFirstTwo)
range(checkLastTwo)
```
Confirmed, supposed hours only take values from 10 to 23, and minutes from 0 to 59. Let's continue checking the observations with one or two digits.
```{r}
# Let's check now the observations with one or two digits.
checkOneTwo <- crimeset$TIME.OCC[which(crimeset$TIME.OCC < 100)]
range(checkOneTwo)
```
Also ok. Lastly, let's check the observations with three digits.
```{r}
# Lastly let's check the observations with three digits.
checkFirstThree <- crimeset$TIME.OCC[which(crimeset$TIME.OCC < 1000)]%/%100
checkLastThree <- crimeset$TIME.OCC[which(crimeset$TIME.OCC < 1000)]%%100
range(checkFirstThree)
range(checkLastThree)
```
Also ok, so it's confirmed, 1 means 00:01 and 2359 means 23:59.

#### **5) Variable AREA:**

```{r}
sort(unique(crimeset$AREA))
```
Areas are codified with a number from 1 to 21, so let's transform them into factor.
```{r}
crimeset$AREA <- factor(crimeset$AREA)

# Relative frequencies barplot
ggplot(as.data.frame(round(table(crimeset$AREA)/length(crimeset$AREA), 2)),
       aes(x = Var1, y = Freq)) + geom_bar(stat = "identity", color = "indianred3", fill = "indianred3") +
  labs(title = "Relative frequencies of variable area", x = "area code") + theme_classic() +
  geom_text(aes(label = Freq), hjust = 1.3) + 
  theme(plot.title = element_text(hjust = 0.5, face = "bold")) + coord_flip()
```

#### **6) Variable AREA.NAME:**

```{r}
head(crimeset$AREA.NAME)
length(unique(crimeset$AREA.NAME))
```
21 unique values, those are the corresponding names of the previous areas. Let's see to which code corrresponds each area name.
```{r}
# Auxiliar dataframe to present the area codes and the area names
auxSet <- data.frame(AREA = numeric(21), AREA_NAME = character(21))
for (i in 1:21) {
  auxSet[i, ] <- head(crimeset[which(crimeset$AREA == i), c("AREA", "AREA.NAME")], 1)
}
auxSet
```
Let's add the area names as the levels of the area variable.
```{r}
levels(crimeset$AREA) <- auxSet$AREA_NAME
head(crimeset$AREA, 7)
```

#### **7) Variable Rpt.Dist.No:**

```{r}
head(crimeset$Rpt.Dist.No)
range(crimeset$Rpt.Dist.No)
```
This variable takes integer values from 101 to 2199.
```{r}
length(unique(crimeset$Rpt.Dist.No))
```
It takes 1209 different values, these are the ten most repeated ones.
```{r}
head(sort(table(crimeset$Rpt.Dist.No), decreasing = T), 10)
```
It's really difficult to deduct the meaning of this variable.

#### **8) Variable Rpt.Dist.No:**

```{r}
unique(crimeset$Part.1.2)
```
It's a dichotomous variable, so let's transform it into a factor.
```{r}
crimeset$Part.1.2 <- factor(crimeset$Part.1.2)

# Relative frequencies barplot
ggplot(as.data.frame(round(table(crimeset$Part.1.2)/length(crimeset$Part.1.2), 2)),
       aes(x = Var1, y = Freq)) + geom_bar(stat = "identity", color = "indianred3", fill = "indianred3") +
  labs(title = "Relative frequencies of variable Part.1.2", x = "Part.1.2") + theme_classic() +
  geom_text(aes(label = Freq), vjust = -0.3) + 
  theme(plot.title = element_text(hjust = 0.5, face = "bold"))
```

The category 1 it's observed in 59 % of the observations and the category 2 in the resting 41 %.

#### **9) Variable Crm.Cd:**


```{r}
head(crimeset$Crm.Cd)
range(crimeset$Crm.Cd)
typeof(crimeset$Crm.Cd)
length(unique(crimeset$Crm.Cd))
```
Again we have an integer variable of difficult interpretation. It takes values in the range 110 to 956. It takes only 139 different values, so it could be a categorical variable. The most repeated are:
```{r}
head(sort(table(crimeset$Crm.Cd), decreasing = T), 10)
```

#### **10) Variable Crm.Cd.Desc:**

```{r}
head(crimeset$Crm.Cd.Desc)
length(unique(crimeset$Crm.Cd.Desc))
```
139 unique different values, this variable accounts the meaning of the previous one. Let's print the 20 first crime codes and their names.
```{r}
# Let's create a dataframe with the crimes codes and their name.
crimeNames <- data.frame(code = numeric(139), crime = character(139))
rowIndex <- 1 # Auxiliar variable for selecting the row
for (i in sort(unique(crimeset$Crm.Cd))) {
  crimeNames[rowIndex, ] <- head(crimeset[which(crimeset$Crm.Cd == i), c("Crm.Cd", "Crm.Cd.Desc")], 1)
  rowIndex <- rowIndex + 1
}

# Result
head(crimeNames, 20)
```
Now that its confirmed that the previous variable was a categorical one, let's create a new variable with the previous one as a factor and assign it the correct level names.
```{r}
crimeset$crm_cd <- factor(crimeset$Crm.Cd)
head(crimeset$crm_cd)
levels(crimeset$crm_cd) <- crimeNames$crime
head(crimeset$crm_cd, 7)
```
Let's view now the 20 most frequent types of crimes registered.
```{r}
data.frame(crime = names(head(sort(table(crimeset$crm_cd), decreasing = T), 20)), 
           frequency = as.vector(head(sort(table(crimeset$crm_cd), decreasing = T), 20)), 
           rel_frequency = round(100*as.vector(head(sort(table(crimeset$crm_cd),
                                                         decreasing = T), 20))/955339, 2))
```
As can be seen the most observed crime is vehicles steal.

#### **11) Variable Mocodes:**

```{r}
head(crimeset$Mocodes)
length(unique(crimeset$Mocodes))
```
Character variable that accounts for some kind of code or id. Only has 307532 unique values.
```{r}
head(sort(table(crimeset$Mocodes), decreasing = T), 7)
```
The most repeated value is the empty string "", so this could mean that there are 136675 missing values in the variable, and it would be convenient to modify them into NA.

#### **12) Variable Vict.Age:**

By its name, this variable must contain the victim's age.
```{r}
range(crimeset$Vict.Age)
```
The range takes negative values, so we must decide if put them as missing or simply has positive. Also the maximum age looks pretty high. Let's check how many ages are negative and how many ages are near 120.
```{r}
# How many negative values are in the variable?
length(which(crimeset$Vict.Age < 0))
```
124 are negative values. Let's see their frequencies.
```{r}
table(crimeset$Vict.Age[which(crimeset$Vict.Age <= 0)])
```
There are 245607 zeros in the variable, over a total of 955339 observations, that would mean that 26 % of the victims in the dataset are infants of less than one year of age, which seems a non-realistic cipher. Let's observe what kind of crimes are this observations associated with.
```{r}
data.frame(crime = names(head(sort(table(crimeset$crm_cd[which(crimeset$Vict.Age == 0)]), decreasing = T), 20)), 
           frequency = as.vector(head(sort(table(crimeset$crm_cd[which(crimeset$Vict.Age == 0)]), decreasing = T), 20)))
```
As can be seen in the table above, the crimes reported for age zero are hardly associated with persons of that age. Let's see what is the other than zero minimum age observed and contrast its results with these.
```{r}
# Ages registered that are less than 10 years.
sort(unique(crimeset$Vict.Age[which(crimeset$Vict.Age < 10)]))
length(which(crimeset$Vict.Age == 2))
100*length(which(crimeset$Vict.Age == 2))/955339
```
The next minimum age is 2 which has 426 observations in this dataset, that's a 0.045 %, which sounds like a more realistic number than the one obtained for age 0.
```{r}
data.frame(crime = names(head(sort(table(crimeset$crm_cd[which(crimeset$Vict.Age == 2)]), decreasing = T), 20)), 
           frequency = as.vector(head(sort(table(crimeset$crm_cd[which(crimeset$Vict.Age == 2)]), decreasing = T), 20)))
```
Also now the most observed crimes are ones that sound like directly or at least can indirectly be associated with children, so effectively let's replace all the zero ages with NA's. For the case of the negative values, it could be that simply were put to negative by mistake, and they only need to be modified as positve. To look for more insights that supports this view, let's observe what associated crimes they have.
```{r}
data.frame(crime = names(head(sort(table(crimeset$crm_cd[which(crimeset$Vict.Age < 0)]), decreasing = T), 26)), 
           frequency = as.vector(head(sort(table(crimeset$crm_cd[which(crimeset$Vict.Age < 0)]), decreasing = T), 26)))
```
Looking at the mentioned frequencies it doesn't seems clear that this observations refer to child related crimes, so also let's replace this observations with NA values.
```{r}
# Replace all negative and zero ages with NA's.
crimeset$Vict.Age <- replace(crimeset$Vict.Age, which(crimeset$Vict.Age <= 0), NA)
range(crimeset$Vict.Age, na.rm = T)
```
New ranges goes from 2 to 120.
Now let's explore how many values are near its maximum, 120 years.
```{r}
table(crimeset$Vict.Age[which(crimeset$Vict.Age > 90)])
```
Only one observation takes the value 120 and the nearest one to 120 is 99. Assuming that the data comes from the United States, if we search about the eldest people registered there, can be seen that until this day, nobody has passed from 119 years of age. So definetly, let's assume that this age also must be replaced with a missing value.
```{r}
# Replace the value that takes 120 by an NA.
crimeset$Vict.Age <- replace(crimeset$Vict.Age, which(crimeset$Vict.Age == 120), NA)
# New range
range(crimeset$Vict.Age, na.rm = T)
summary(crimeset$Vict.Age)

# Histogram
ggplot(crimeset, aes(x = Vict.Age)) + geom_histogram(color = "white", fill = "indianred3", binwidth = 5) +
  theme_light() + labs(title = "Histogram of variable Vict.Age", x = "victim's age") +
  theme(plot.title = element_text(hjust = 0.5, face = "bold"))

# Boxplot
ggplot(crimeset, aes(x = Vict.Age)) + geom_boxplot(color = "indianred3") + coord_flip() +
  labs(title = "Boxplot of variable Vict.Age", x = "value") + 
  theme(plot.title = element_text(hjust = 0.5, face = "bold"))
```

Now the minimum value is 2, the maximum value is 99 and the median is 37. Missing values now are around the 26 % of the variable.

#### **13) Variable Vict.Sex**

```{r}
head(crimeset$Vict.Sex)
```
This variable accounts for the victim's sex, currently is a character variable, so it needs to be transformed into a factor, but first let's view which values it takes.
```{r}
table(crimeset$Vict.Sex)
```
We see that 130045 observations take the empty character "", 1 observation take the character "-", 109 the letter "H" and 89366 the letter "X", let's assume that all this observations are missing values, so let's replace them with NA's.
```{r}
crimeset$Vict.Sex <- replace(crimeset$Vict.Sex, which(crimeset$Vict.Sex == ""), NA)
crimeset$Vict.Sex <- replace(crimeset$Vict.Sex, which(crimeset$Vict.Sex == "-"), NA)
crimeset$Vict.Sex <- replace(crimeset$Vict.Sex, which(crimeset$Vict.Sex == "H"), NA)
crimeset$Vict.Sex <- replace(crimeset$Vict.Sex, which(crimeset$Vict.Sex == "X"), NA)
round(100*sum(is.na(crimeset$Vict.Sex))/955339, 2)

# Relative frequencies barplot
ggplot(as.data.frame(round(table(crimeset$Vict.Sex)/length(crimeset$Vict.Sex), 2)),
       aes(x = Var1, y = Freq)) + geom_bar(stat = "identity", color = "indianred3", fill = "indianred3") +
  labs(title = "Relative frequencies of variable victim's sex", x = "victim's sex") + theme_classic() +
  geom_text(aes(label = Freq), vjust = -0.5) + 
  theme(plot.title = element_text(hjust = 0.5, face = "bold"))
```

There are 41 % of males and 36 % of females in the observations. Now 22.98 % of the observations are missing values.

#### **14) Variable Premis.Cd**

