---
title: "**CRIME DATA ANALYSIS**"
output: html_document
---

### **MAIN OBJECTIVE**

The objective of this notebook is to explore the crime dataset found in kaggle at: <https://www.kaggle.com/datasets/haseefalam/crime-dataset/data?select=Crime_Data_from_2020_to_Present.csv>.

### **NECESSARY LIBRARIES**

Those are the libraries that are going to be used across this notebook.

```{r}
library(ggplot2)
library(lubridate)
```

### **DATA LOAD**

```{r}
setwd("C://Users//riosa//documents//datasets//crime_dataset") # sets the directory to the specified path
crimeset <- read.csv("Crime_Data_from_2020_to_Present.csv") # loads the dataset
```

### **EXPLORATORY DATA ANALYSIS**

```{r}
dim(crimeset)
```
The dataset counts with 955339 observations and 28 variables.
Let's check the variable names.
```{r}
names(crimeset)
```
There are missing values in some variable?
```{r}
missing_table <- as.matrix(colSums(apply(X = crimeset, MARGIN = 2, FUN = is.na)))
missing_table
```
The percentages of missing values in the different variables are:
```{r}
for (i in 1:dim(missing_table)[1]) {
  if (missing_table[i, 1] != 0) {
    cat("Percentage of missings of variable ", names(missing_table[i, ]), ":", round(100*missing_table[i, 1]/955339, 2), "%\n")
  }
}
```
Variables Crm.Cd.2, Crm.Cd.3 and Crm.Cd.4 have more than a 90 % of missing values, so they are hardly useful for modelling.

Let's explore now the contents of each variable:

#### **1) Variable DR_NO:**

```{r}
head(crimeset$DR_NO)
length(unique(crimeset$DR_NO))
length(unique(crimeset$DR_NO)) == dim(crimeset)[1]
```
This variable contains a unique identifier for each observation, so this variable will not be useful for modelling.

#### **2) Variable Date.Rptd:**

```{r}
head(crimeset$Date.Rptd)
```
This variable contains dates and times. The first thing we will do is to check how many different hours the variable takes.
```{r}
# Let's see how to proceed for one case...
paste(unlist(strsplit(crimeset$Date.Rptd[1], " "))[2:3], collapse = " ")

# Let's create an auxiliar funciton to perform the above task
getHour <- function(value) {
  # Input a character string like "03/01/2020 12:00:00 AM"
  # Return a character string like "12:00:00 AM"
  # Input is assumed to be passed correctly by the user
  return(paste(unlist(strsplit(value, " "))[2:3], collapse = " "))
}

# Testing...
getHour(crimeset$Date.Rptd[1])
getHour(crimeset$Date.Rptd[2])
getHour(crimeset$Date.Rptd[3])

# Let's store all the hours in a vector
checkHours <- vapply(crimeset$Date.Rptd, getHour, character(1))
# Vector obtained
head(checkHours)
# Let's check how many different values it takes...
length(unique(checkHours))
```
There is only one hour that repeats across all the rows for this variable, 12:00:00 AM, so let's remove them and only remain with the date. For doing this, let's add a new variable to the dataset.
```{r}
# Let's create an auxiliar function to do the process...
getDate <- function(value) {
  # Input: a character string with the format: "03/01/2020 12:00:00 AM"
  # Return: a character type with the format: "03/01/2020"
  # Input is assumed to be correctly passed by the user
  return(unlist(strsplit(value, " "))[1])
}

# Now let's create a new variable withe the dates in the dataset crimeset
crimeset$rptd_date <-  vapply(crimeset$Date.Rptd, getDate, character(1))
crimeset$rptd_date <- as.Date(crimeset$rptd_date, format = "%m/%d/%Y")

# Result
head(crimeset$rptd_date)
range(crimeset$rptd_date)
```
The range of the dates goes from the first of January of 2020 to the 24th of June of 2024.
```{r}
length(unique(crimeset$rptd_date))
```
There are only 1637 different dates for 955339 observations. Let's see which ones repeat the most.
```{r}
# Ten most repeated dates
head(sort(table(crimeset$rptd_date), decreasing = T), 10)
```
Let's check now the frequencies of each year.
```{r}
# Frequencies of the years
table(year(crimeset$rptd_date))/sum(table(year(crimeset$rptd_date)))
```
The year with more observations is 2022 with the 24.61 %, followed by 2023 with the 24.56 %. The year with less observations is 2024 which only accounts for the 8.86 % of the observations, although there are observations until the mid of 2024.

#### **3) Variable DATE.OCC:**

```{r}
head(crimeset$DATE.OCC)
```
Let's proceed like in the previous variable, first let's check if all the observations take
the same hour.
```{r}
# Let's store all the hours in a vector
checkHours2 <- vapply(crimeset$DATE.OCC, getHour, character(1))
head(checkHours2)
length(unique(checkHours2))
```
Again, there is the same hour for all the rows, so let's save in a new variable only the dates.
```{r}
crimeset$date_occ <-  vapply(crimeset$DATE.OCC, getDate, character(1))
crimeset$date_occ <- as.Date(crimeset$date_occ, format = "%m/%d/%Y")
head(crimeset$date_occ)
range(crimeset$date_occ)
```
The range of dates is the same of the previous variable.
```{r}
head(crimeset$rptd_date)
head(crimeset$date_occ)
```
It seems that the dates of the variable rptd_date can be always older or equeal than the dates of the variable date_occ, let's check it.
```{r}
checkPositive <- crimeset$rptd_date - crimeset$date_occ
head(checkPositive, 10)
length(which(checkPositive >= 0))
```
Confirmed, it's true for all rows.

#### **4) Variable TIME.OCC:**

```{r}
head(crimeset$TIME.OCC)
summary(crimeset$TIME.OCC)
head(sort(table(crimeset$TIME.OCC), decreasing = T), 10)
```
It's not clear the meaning and the format of this variable, by its name, it's supposed to contain hours and maybe minutes, but its format is of integer values from 1 to 2359. Values are repeated many times so maybe they can not take any value from 1 to 2359. It could be that 1 means 00:01 and 2359 means 23:59. For checking if this is possible let's test that in all the obervations with four digits the first two never take values bigger than 23, and that the last two digits never take values bigger than 59.
```{r}
# Let's check first only the observations of four digits.
checkFirstTwo <- crimeset$TIME.OCC[which(crimeset$TIME.OCC >= 1000)]%/%100
checkLastTwo <- crimeset$TIME.OCC[which(crimeset$TIME.OCC >= 1000)]%%100
range(checkFirstTwo)
range(checkLastTwo)
```
Confirmed, supposed hours only take values from 10 to 23, and minutes from 0 to 59. Let's continue checking the observations with one or two digits.
```{r}
# Let's check now the observations with one or two digits.
checkOneTwo <- crimeset$TIME.OCC[which(crimeset$TIME.OCC < 100)]
range(checkOneTwo)
```
Also ok. Lastly, let's check the observations with three digits.
```{r}
# Lastly let's check the observations with three digits.
checkFirstThree <- crimeset$TIME.OCC[which(crimeset$TIME.OCC < 1000)]%/%100
checkLastThree <- crimeset$TIME.OCC[which(crimeset$TIME.OCC < 1000)]%%100
range(checkFirstThree)
range(checkLastThree)
```
Also ok, so it's confirmed, 1 means 00:01 and 2359 means 23:59.

#### **5) Variable AREA:**

```{r}
sort(unique(crimeset$AREA))
```
Areas are codified with a number from 1 to 21, so let's transform them into factor.
```{r}
crimeset$AREA <- factor(crimeset$AREA)

# Relative frequencies barplot
ggplot(as.data.frame(round(table(crimeset$AREA)/length(crimeset$AREA), 2)),
       aes(x = Var1, y = Freq)) + geom_bar(stat = "identity", color = "indianred3", fill = "indianred3") +
  labs(title = "Relative frequencies of variable area", x = "area code") + theme_classic() +
  geom_text(aes(label = Freq), hjust = 1.3) + 
  theme(plot.title = element_text(hjust = 0.5, face = "bold")) + coord_flip()
```

#### **6) Variable AREA.NAME:**

```{r}
head(crimeset$AREA.NAME)
length(unique(crimeset$AREA.NAME))
```
21 unique values, those are the corresponding names of the previous areas. Let's see to which code corrresponds each area name.
```{r}
# Auxiliar dataframe to present the area codes and the area names
auxSet <- data.frame(AREA = numeric(21), AREA_NAME = character(21))
for (i in 1:21) {
  auxSet[i, ] <- head(crimeset[which(crimeset$AREA == i), c("AREA", "AREA.NAME")], 1)
}
auxSet
```
Let's add the area names as the levels of the area variable.
```{r}
levels(crimeset$AREA) <- auxSet$AREA_NAME
head(crimeset$AREA, 7)
```

#### **7) Variable Rpt.Dist.No:**

```{r}
head(crimeset$Rpt.Dist.No)
range(crimeset$Rpt.Dist.No)
```
This variable takes integer values from 101 to 2199.
```{r}
length(unique(crimeset$Rpt.Dist.No))
```
It takes 1209 different values, these are the ten most repeated ones.
```{r}
head(sort(table(crimeset$Rpt.Dist.No), decreasing = T), 10)
```
It's really difficult to deduct the meaning of this variable.

#### **8) Variable Rpt.Dist.No:**

```{r}
unique(crimeset$Part.1.2)
```
It's a dichotomous variable, so let's transform it into a factor.
```{r}
crimeset$Part.1.2 <- factor(crimeset$Part.1.2)

# Relative frequencies barplot
ggplot(as.data.frame(round(table(crimeset$Part.1.2)/length(crimeset$Part.1.2), 2)),
       aes(x = Var1, y = Freq)) + geom_bar(stat = "identity", color = "indianred3", fill = "indianred3") +
  labs(title = "Relative frequencies of variable Part.1.2", x = "Part.1.2") + theme_classic() +
  geom_text(aes(label = Freq), vjust = -0.3) + 
  theme(plot.title = element_text(hjust = 0.5, face = "bold"))
```

The category 1 it's observed in 59 % of the observations and the category 2 in the resting 41 %.

#### **9) Variable Crm.Cd:**


```{r}
head(crimeset$Crm.Cd)
range(crimeset$Crm.Cd)
typeof(crimeset$Crm.Cd)
length(unique(crimeset$Crm.Cd))
```
Again we have an integer variable of difficult interpretation. It takes values in the range 110 to 956. It takes only 139 different values, so it could be a categorical variable. The most repeated are:
```{r}
head(sort(table(crimeset$Crm.Cd), decreasing = T), 10)
```

#### **10) Variable Crm.Cd.Desc:**

```{r}
head(crimeset$Crm.Cd.Desc)
length(unique(crimeset$Crm.Cd.Desc))
```
139 unique different values, this variable accounts the meaning of the previous one. Let's print the 20 first crime codes and their names.
```{r}
# Let's create a dataframe with the crimes codes and their name.
crimeNames <- data.frame(code = numeric(139), crime = character(139))
rowIndex <- 1 # Auxiliar variable for selecting the row
for (i in sort(unique(crimeset$Crm.Cd))) {
  crimeNames[rowIndex, ] <- head(crimeset[which(crimeset$Crm.Cd == i), c("Crm.Cd", "Crm.Cd.Desc")], 1)
  rowIndex <- rowIndex + 1
}

# Result
head(crimeNames, 20)
```
Now that its confirmed that the previous variable was a categorical one, let's create a new variable with the previous one as a factor and assign it the correct level names.
```{r}
crimeset$crm_cd <- factor(crimeset$Crm.Cd)
head(crimeset$crm_cd)
levels(crimeset$crm_cd) <- crimeNames$crime
head(crimeset$crm_cd, 7)
```
Let's view now the 20 most frequent types of crimes registered.
```{r}
data.frame(crime = names(head(sort(table(crimeset$crm_cd), decreasing = T), 20)), 
           frequency = as.vector(head(sort(table(crimeset$crm_cd), decreasing = T), 20)), 
           rel_frequency = round(100*as.vector(head(sort(table(crimeset$crm_cd),
                                                         decreasing = T), 20))/955339, 2))
```
As can be seen the most observed crime is vehicles steal.

#### **11) Variable Mocodes:**

```{r}
head(crimeset$Mocodes)
length(unique(crimeset$Mocodes))
```
Character variable that accounts for some kind of code or id. Only has 307532 unique values.
```{r}
head(sort(table(crimeset$Mocodes), decreasing = T), 7)
```
The most repeated value is the empty string "", so this could mean that there are 136675 missing values in the variable, and it would be convenient to modify them into NA.

#### **12) Variable Vict.Age:**

By its name, this variable must contain the victim's age.
```{r}
range(crimeset$Vict.Age)
```
The range takes negative values, so we must decide if put them as missing or simply has positive. Also the maximum age looks pretty high. Let's check how many ages are negative and how many ages are near 120.
```{r}
# How many negative values are in the variable?
length(which(crimeset$Vict.Age < 0))
```
124 are negative values. Let's see their frequencies.
```{r}
table(crimeset$Vict.Age[which(crimeset$Vict.Age <= 0)])
```
There are 245607 zeros in the variable, over a total of 955339 observations, that would mean that 26 % of the victims in the dataset are infants of less than one year of age, which seems a non-realistic cipher. Let's observe what kind of crimes are this observations associated with.
```{r}
data.frame(crime = names(head(sort(table(crimeset$crm_cd[which(crimeset$Vict.Age == 0)]), decreasing = T), 20)), 
           frequency = as.vector(head(sort(table(crimeset$crm_cd[which(crimeset$Vict.Age == 0)]), decreasing = T), 20)))
```
As can be seen in the table above, the crimes reported for age zero are hardly associated with persons of that age. Let's see what is the other than zero minimum age observed and contrast its results with these.
```{r}
# Ages registered that are less than 10 years.
sort(unique(crimeset$Vict.Age[which(crimeset$Vict.Age < 10)]))
length(which(crimeset$Vict.Age == 2))
100*length(which(crimeset$Vict.Age == 2))/955339
```
The next minimum age is 2 which has 426 observations in this dataset, that's a 0.045 %, which sounds like a more realistic number than the one obtained for age 0.
```{r}
data.frame(crime = names(head(sort(table(crimeset$crm_cd[which(crimeset$Vict.Age == 2)]), decreasing = T), 20)), 
           frequency = as.vector(head(sort(table(crimeset$crm_cd[which(crimeset$Vict.Age == 2)]), decreasing = T), 20)))
```
Also now the most observed crimes are ones that sound like directly or at least can indirectly be associated with children, so effectively let's replace all the zero ages with NAs. For the case of the negative values, it could be that simply were put to negative by mistake, and they only need to be modified as positve. To look for more insights that supports this view, let's observe what associated crimes they have.
```{r}
data.frame(crime = names(head(sort(table(crimeset$crm_cd[which(crimeset$Vict.Age < 0)]), decreasing = T), 26)), 
           frequency = as.vector(head(sort(table(crimeset$crm_cd[which(crimeset$Vict.Age < 0)]), decreasing = T), 26)))
```
Looking at the mentioned frequencies it doesn't seems clear that this observations refer to child related crimes, so also let's replace this observations with NA values.
```{r}
# Replace all negative and zero ages with NAs.
crimeset$Vict.Age <- replace(crimeset$Vict.Age, which(crimeset$Vict.Age <= 0), NA)
range(crimeset$Vict.Age, na.rm = T)
```
New ranges goes from 2 to 120.
Now let's explore how many values are near its maximum, 120 years.
```{r}
table(crimeset$Vict.Age[which(crimeset$Vict.Age > 90)])
```
Only one observation takes the value 120 and the nearest one to 120 is 99. Assuming that the data comes from the United States, if we search about the eldest people registered there, can be seen that until this day, nobody has passed from 119 years of age. So definetly, let's assume that this age also must be replaced with a missing value.
```{r}
# Replace the value that takes 120 by an NA.
crimeset$Vict.Age <- replace(crimeset$Vict.Age, which(crimeset$Vict.Age == 120), NA)
# New range
range(crimeset$Vict.Age, na.rm = T)
summary(crimeset$Vict.Age)

# Histogram
ggplot(crimeset, aes(x = Vict.Age)) + geom_histogram(color = "white", fill = "indianred3", binwidth = 5) +
  theme_light() + labs(title = "Histogram of variable Vict.Age", x = "victim's age") +
  theme(plot.title = element_text(hjust = 0.5, face = "bold"))

# Boxplot
ggplot(crimeset, aes(x = Vict.Age)) + geom_boxplot(color = "indianred3") + coord_flip() +
  labs(title = "Boxplot of variable Vict.Age", x = "value") + 
  theme(plot.title = element_text(hjust = 0.5, face = "bold"))
```

Now the minimum value is 2, the maximum value is 99 and the median is 37. Missing values now are around the 26 % of the variable.

#### **13) Variable Vict.Sex**

```{r}
head(crimeset$Vict.Sex)
```
This variable accounts for the victim's sex, currently is a character variable, so it needs to be transformed into a factor, but first let's view which values it takes.
```{r}
table(crimeset$Vict.Sex)
```
We see that 130045 observations take the empty character "", 1 observation take the character "-", 109 the letter "H" and 89366 the letter "X", let's assume that all this observations are missing values, so let's replace them with NAs.
```{r}
crimeset$Vict.Sex <- replace(crimeset$Vict.Sex, which(crimeset$Vict.Sex == ""), NA)
crimeset$Vict.Sex <- replace(crimeset$Vict.Sex, which(crimeset$Vict.Sex == "-"), NA)
crimeset$Vict.Sex <- replace(crimeset$Vict.Sex, which(crimeset$Vict.Sex == "H"), NA)
crimeset$Vict.Sex <- replace(crimeset$Vict.Sex, which(crimeset$Vict.Sex == "X"), NA)
round(100*sum(is.na(crimeset$Vict.Sex))/955339, 2)

# Relative frequencies barplot
ggplot(as.data.frame(round(table(crimeset$Vict.Sex)/length(crimeset$Vict.Sex), 2)),
       aes(x = Var1, y = Freq)) + geom_bar(stat = "identity", color = "indianred3", fill = "indianred3") +
  labs(title = "Relative frequencies of variable victim's sex", x = "victim's sex") + theme_classic() +
  geom_text(aes(label = Freq), vjust = -0.5) + 
  theme(plot.title = element_text(hjust = 0.5, face = "bold"))
```

There are 41 % of males and 36 % of females in the observations. Now 22.98 % of the observations are missing values.

#### **14) Variable Premis.Cd**

```{r}
head(crimeset$Premis.Cd)
sum(is.na(crimeset$Premis.Cd))
range(crimeset$Premis.Cd, na.rm = T)
length(unique(crimeset$Premis.Cd))
```
Again it seems that we have a codified categorical varaible, this time with 12 missing values and 315 different categories.

#### **15) Variable Premis.Desc**

```{r}
head(crimeset$Premis.Desc)
```
This variable looks like the description of the categories of the previous one.
```{r}
length(unique(crimeset$Premis.Desc))
sum(is.na(crimeset$Premis.Desc))
```
It has 307 unique different values vs. the 315 unique different values the previous one had, also this variable has no missing values. So, it seems that can't be a perfect match between codified categories and descriptions. Let's check if there are empty categories in the variable.
```{r}
length(which(crimeset$Premis.Desc == ""))
```
There are 569 observations which its value is "", so these observations could be replaced with NAs and therefore treated as missing. Let's check what codes are associated with the empty description.
```{r}
unique(crimeset$Premis.Cd[which(crimeset$Premis.Desc == "")])
table(crimeset$Premis.Cd[which(crimeset$Premis.Desc == "")])
569 - sum(table(crimeset$Premis.Cd[which(crimeset$Premis.Desc == "")]))
```
There are 8 different codes that take an empty description, among them are the 12 observations that had a missing code. Let's check if this codes only take the empty description.
```{r}
for (i in unique(crimeset$Premis.Cd[which(crimeset$Premis.Desc == "")])[-4]) {
  cat("code:", i, "\n")
  print(table(crimeset$Premis.Desc[which(crimeset$Premis.Cd == i)]))
  cat("--------------------------\n")
}
```
All the eight codes (the ones with NAs included) only take the empty description, so we can replace all this observations with NAs.
```{r}
# Replace all the empty descriptions with NA values.
crimeset$Premis.Desc <- replace(crimeset$Premis.Desc, which(crimeset$Premis.Desc == ""), NA)
# Checking...
sum(is.na(crimeset$Premis.Desc))
```
Now let's take a look at some of the codes and its descriptions.
```{r}
# Let's create a dataframe with the different descriptions and their associated codes.
#
# Dataframe initialization
premisFrame <- data.frame(code = numeric(306), description = character(306))
rowIndex <- 1
for (i in unique(crimeset$Premis.Desc)[-125]) {
  premisFrame[rowIndex, ] <- head(crimeset[which(crimeset$Premis.Desc == i), c("Premis.Cd", "Premis.Desc")], 1)
  rowIndex <- rowIndex  + 1
}

# Sample result
head(premisFrame, 20)
```
Now let's convert the variable into a factor.
```{r}
# Convert the variable into a factor
crimeset$Premis.Desc <- factor(crimeset$Premis.Desc)

```
Let's observe the 10 most frequent scenarios of the registered crimes.
```{r}
# Most observerd ones
head(sort(table(crimeset$Premis.Desc), decreasing = T), 10)
```
As can be seen, the four most repeated crime scenarios are the street, homes and parking lots.

#### **16) Variable Weapon.Used.Cd**

Again it seems that we have a numerically codified categorical variable.
```{r}
head(crimeset$Weapon.Used.Cd)
sum(is.na(crimeset$Weapon.Used.Cd))
100*sum(is.na(crimeset$Weapon.Used.Cd))/dim(crimeset)[1]
```
66 % of the observation are missing values.
```{r}
range(crimeset$Weapon.Used.Cd, na.rm = T)
```
The range of the codes goes from 101 to 516. The codes most repeated are:
```{r}
head(sort(table(crimeset$Weapon.Used.Cd), decreasing = T))
```

#### **17) Variable Weapon.Desc**

```{r}
head(crimeset$Weapon.Desc)
sum(is.na(crimeset$Weapon.Desc))
length(which(crimeset$Weapon.Desc == ""))
```
Again seems that the missing values are codified as empty strings, to confirm it, let's check the coincidence between the indexes of the previous variable that took the value NA and the indexes of this variable that take the empty string value.
```{r}
# Checking that the rows with NAs of the previous variable and the rows with "" of this one are
# the same.
rowsCode <- which(is.na(crimeset$Weapon.Used.Cd))
rowsDesc <- which(crimeset$Weapon.Desc == "")
length(rowsCode) == length(rowsDesc)
# Lengths are the same, ok
sum(rowsCode == rowsDesc)
```
Coincidence between both indexes is confirmed, so let's replace the empty strings with NAs.
```{r}
# Replacing the empty strings with NAs.
crimeset$Weapon.Desc <- replace(crimeset$Weapon.Desc, which(crimeset$Weapon.Desc == ""), NA)
length(unique(crimeset$Weapon.Desc))
```
Now the variable has 80 different values among which are the ones with NAs. Let's transform it into a factor variable.
```{r}
crimeset$Weapon.Desc <- factor(crimeset$Weapon.Desc)
```
Let's observe now which are the 20 weapons most observed in the dataset.
```{r}
data.frame(weapon = names(head(sort(table(crimeset$Weapon.Desc), decreasing = T), 20)), 
           frequency = as.vector(head(sort(table(crimeset$Weapon.Desc), decreasing = T), 20)))
```
As can be seen, the weapon most used was strong-arm. Let's explore now which types of crimes are more observed for the categories strong-arm, verbal threat and vehicle.
```{r}
for (i in c("STRONG-ARM (HANDS, FIST, FEET OR BODILY FORCE)", "VERBAL THREAT", "VEHICLE")) {
  cat("Weapon:", i, "\n\n")
  print.data.frame(data.frame(crime = names(head(sort(table(crimeset$Crm.Cd.Desc[which(crimeset$Weapon.Desc == i)]), decreasing = T), 10)), frequency = as.vector(head(sort(table(crimeset$Crm.Cd.Desc[which(crimeset$Weapon.Desc == i)]), decreasing = T), 10))), row.names = F)
  cat("-----------------------------------------------------------------------\n")
}
```

#### **18) Variable Status**

```{r}
table(crimeset$Status)
```
This variable takes eight different values among which is the empty string "", so let's replace this observation with a NA. Also let's transform the variable into the more appropriate type factor. Anyway, without more information will be diffictult to advert the meaning of the elements of this variable.
```{r}
# Replace empty string with NA
crimeset$Status <- replace(crimeset$Status, which(crimeset$Status == ""), NA)
sum(is.na(crimeset$Status))
# Transform into factor
crimeset$Status <- factor(crimeset$Status)
```

#### **19) Variable Status.Desc**

```{r}
head(crimeset$Status.Desc)
```
This varaible seems to contain the descriptions of the previous one, let's see which values it takes.
```{r}
unique(crimeset$Status.Desc)
```
The different categories seems to mean "Adult Arrested", "Investigation Continues", "Juvenile Arrested", "Adult Other", "Juvenile Other" and "UNKNOWN", so this last one should be the category associated with the NA value of the previous variable, let's check it.
```{r}
crimeset$Status.Desc[which(is.na(crimeset$Status))]
```
Ok, it's confirmed, so this observation can be replaced by an NA in this variable also.
Now let's see if more observations take the value "UNK" and if it's affirmative, proceed to look for their corresponding value in the previous variable.
```{r}
# How many observations take the value "UNK"?
length(which(crimeset$Status.Desc == "UNK"))
# 6 observations take the value "UNK"
data.frame(status = crimeset$Status[which(crimeset$Status.Desc == "UNK")], 
           status_desc = rep("UNK", 6))
```
So we can see that there is one more value in the variable status that indicates a missing observation, that's the category "CC". This time, instead of transforming directly the variable status.desc, let's create a new one and put it as a factor.
```{r}
crimeset$status_desc <- replace(crimeset$Status.Desc, which(crimeset$Status.Desc == "UNK"), NA)
sum(is.na(crimeset$status_desc))
crimeset$status_desc <- factor(crimeset$status_desc)

# Absolute frequencies barplot
ggplot(as.data.frame(table(crimeset$status_desc)),
       aes(x = Var1, y = Freq)) + geom_bar(stat = "identity", color = "indianred3", fill = "indianred3") +
  labs(title = "Relative frequencies of variable status", x = "status") + theme_classic() +
  geom_text(aes(label = Freq), vjust = -0.3) + 
  theme(plot.title = element_text(hjust = 0.5, face = "bold"))
```

The category "Investigation Continues" is taked by the 80 % of the observations, both adult categories take around the 10 % of the observations and the both juvenile categories take less than the 1 % of the observations.

#### **20) Variables Crm.Cd.1 - 4**

```{r}
range(crimeset$Crm.Cd)
range(crimeset$Crm.Cd.1, na.rm = T)
range(crimeset$Crm.Cd.2, na.rm = T)
range(crimeset$Crm.Cd.3, na.rm = T)
range(crimeset$Crm.Cd.4, na.rm = T)
```
Now we have four variables that contain integer values, although not all the ranges are the same nor are included in the range of the variable Crm.Cd, let's think about this variable and the variable Crm.Cd.Desc as a dictionary to relate codes and their descriptions, and let's take the variables Crm.Cd.1 to Crm.Cd.4 as different crimes charged in relation to one same event.
As has been said, the variables Crm.Cd.1 through Crm.Cd.4 are going to have codes that don't match with the ones in the dictionary variable Crm.Cd, so let's replace all this codes by NAs. To not lose information, let's do this in four new variables. Once this has been done, let's transform these variables into factors and create a function to obtain easily the corresponding description to any code.
```{r}
# Create the new variables with the same information
crimeset$crime1 <- crimeset$Crm.Cd.1
crimeset$crime2 <- crimeset$Crm.Cd.2
crimeset$crime3 <- crimeset$Crm.Cd.3
crimeset$crime4 <- crimeset$Crm.Cd.4

# Replace the non-matching codes by NAs
uniquesZero <- sort(unique(crimeset$Crm.Cd)) # Auxiliar variable with the unique observations of
                                             # Crm.Cd
for (i in c("crime1", "crime2", "crime3", "crime4")) {
  uniquesi <- sort(unique(crimeset[, i])) # Uniques of column i
  dontMatch <- numeric() # Auxiliar variable to store the codes of crimei that don't match any code
  # in Crm.Cd
  for (j in uniquesi) {
    if (length(which(uniquesZero == j)) == 0) {
      dontMatch <- c(dontMatch, j)
    }
  }
  
  # Replace the codes that don't match with NAs
  for (k in dontMatch) {
    crimeset[, i] <- replace(crimeset[, i], which(crimeset[, i] == k), NA)
  }
}

# Transform the four variables into factor
crimeset$crime1 <- factor(crimeset$crime1)
crimeset$crime2 <- factor(crimeset$crime2)
crimeset$crime3 <- factor(crimeset$crime3)
crimeset$crime4 <- factor(crimeset$crime4)

# Create a function that return the description of a given code
getDescription <- function(code) {
  # Receives a positive integer code and if it's one of the values of the 
  # variable Crm.Cd returns the associated description.
  # Example:
  # Input: 510; Return: VEHICLE - STOLEN
  result <- NA # return variable, by default is NA
  checkCode <- unique(crimeset$Crm.Cd.Desc[which(crimeset$Crm.Cd == code)])
  if (length(checkCode) > 0) {
    result <- checkCode
  }
  return(result)
}

# Create a function that returns the code of a given description
getCode <- function(description) {
  # Recieves a character that must match one of the descriptions in Crm.Cd.Desc
  # Example:
  # Input: "VEHICLE - STOLEN"; Return: 510
  result <- NA # variable to be returned, NA by default
  checkDesc <- unique(crimeset$Crm.Cd[which(crimeset$Crm.Cd.Desc == description)])
  if (length(checkDesc) > 0) {
    result <- checkDesc
  }
  return(result)
}
```
Let's observe which values get the variables crime1, crime2 and crime3 when crime4 is not missing.
```{r}
testSet <- crimeset[which(!is.na(crimeset$crime4)),
                    c("crime1", "crime2", "crime3", "crime4")]

# Dataframe with the codes
testSet

testSetDesc <- data.frame(crime1 = character(5), crime2 = character(5), crime3 = character(5), 
                          crime4 = character(5))

for (j in 1:dim(testSetDesc)[2]) {
  testSetDesc[, j] <- vapply(testSet[, j], getDescription, character(1))
}

# Dataframe with the descriptions
testSetDesc
```
So we can see that for example, when the variable crime4 take the value "KIDNAPPING", the variable crime3 took the value "BRANDISH WEAPON", crime2 took the value "ASSAULT WITH DEADLY WEAPON, AGGRAVATED ASSAULT" and crime1 took the value "ROBBERY", so it seems that it was an armed robbery with kidnapping of the workers or house holders.

Let's explore now the ten most observed crimes in the variable crime1.
```{r}
vapply(as.numeric(names((head(sort(table(crimeset$crime1), decreasing = T), 10)))), getDescription, character(1))
```
Let's see which are the weapons used when the crime1 variable takes the value "THEFT OF IDENTITY".
```{r}
head(sort(table(crimeset$Weapon.Desc[which(crimeset$crime1 == getCode("THEFT OF IDENTITY"))]), decreasing = T), 5)
```
Let's also see which values take the other crime variables when crime1 is "THEFT OF IDENTITY".
```{r}
sort(table(crimeset$crime2[which(crimeset$crime1 == getCode("THEFT OF IDENTITY"))]), decreasing = T)
```
It takes thirteen different values, now let's see their descriptions.
```{r}
vapply(head(as.numeric(names(sort(table(crimeset$crime2[which(crimeset$crime1 == getCode("THEFT OF IDENTITY"))]), decreasing = T))), 13), getDescription, character(1))
```
Then the three most observed crimes in the variable crime2 when crime1 is "THEFT OF IDENTITY" are related to document forgery, credit card fraud and plain theft.

#### **21) Variable LOCATION**
